---
title: "Lab 2: Precipitation & Streamflow Relationships"
author: "Kyle Hudson"
id: "1584612"
date: "`r Sys.Date()`"
output:
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r Packages, include=FALSE}
library(tidyverse) #for bloody everything
library(ggeffects) 
```

```{r Input Data, include=FALSE}
HydroData <- read.csv("Lab2Data.csv",
                      header = TRUE,
                      stringsAsFactors = TRUE)

```

Part one
```{r Precipitation Summary, echo=FALSE, message=FALSE, warning=FALSE}
#A table of monthly precipitation
PbyMonth <-
  HydroData %>%
  group_by(Year, NumMon) %>% 
  summarise(Psum = sum(TotalPrecipitation.mm.d., na.rm=T)) %>% #get the sum of each month
  pivot_wider(names_from = NumMon, values_from = Psum) %>% #pivot to make each month a column
  ungroup(.) %>% #undo the grouping now
  mutate(AnnualSum = rowSums(.[2:13])) %>% #get the annual sum in a new column
  mutate(Year = as.character(Year))

#same table, with the mean and SE rows added
Precip_Table <-
  PbyMonth %>% 
  bind_rows(list(summarise_all(., 
                               ~(if( is.numeric(.))
                                 mean(.) 
                                 else 
                                   "Mean Monthly P (mm)"))), 
            (summarise_all(., 
                           ~(if( 
                             is.numeric(.))
                             sd(.)/sqrt(length(.)) 
                             else 
                               "Standard Error")))) %>% #add a standard error row
  mutate(across((c(2:14)), round, 1)) #remember the bracket pattern, it's a pain


Precip_TTest <- #two-sided is default
  t.test(PbyMonth$AnnualSum, 
         conf.level = 0.95)

```

```{r Runoff Summary, include=FALSE}
#A table of monthly runoff 
QbyMonth <-
  HydroData %>%
  group_by(Year, NumMon) %>%
  summarise(Qsum = sum(MeanQmm.day, na.rm=T)) %>% #get the sum of each month
  pivot_wider(names_from = NumMon, values_from = Qsum) %>% #pivot to make each month a column
  ungroup(.) %>% #undo the grouping now
  mutate(AnnualSum = rowSums(.[2:13])) %>% #get the annual sum in a new column
  mutate(Year = as.character(Year))

#same table, with the mean and SE rows added
Runoff_Table <-
  QbyMonth %>% #turn year to a character
  bind_rows(list(summarise_all(., 
                               ~(if(
                                 is.numeric(.))
                                 mean(.) 
                               else 
                                 "Mean Monthly Q (mm)"))), #add a mean row
                 (summarise_all(., 
                               ~(if(
                                 is.numeric(.))
                                 sd(.)/sqrt(length(.)) 
                               else 
                                 "Standard Error")))) %>% #add a standard error row
    mutate(across((c(2:14)), round, 1)) #remember the bracket pattern, it's a pain

Runoff_TTest <- #two-sided is default
  t.test(QbyMonth$AnnualSum, 
         conf.level = 0.95)

```

```{r Manipulate dataframes, include=FALSE}

#Create a dataframe with both precipitation and runoff summed annually
YearData <-
  HydroData %>%
  group_by(Year) %>%
  summarise(RainSum = sum(TotalRainfall.mm.d.),
            SnowSum = sum(TotalSnowfall.mm.deqv..),
            PrecipSum = sum(TotalPrecipitation.mm.d., na.rm=T),
            RunoffSum = sum(MeanQmm.day)) %>%
  mutate(RainPer = (100*RainSum/PrecipSum),
         SnowPer = (100*SnowSum/PrecipSum))

#Create a dataframe with both variables monthly, including standard errors 
MonthData <- 
  t(Precip_Table[c(31,32), ]) %>% #t is transpose
  cbind(t(Runoff_Table[c(31,32), ])) %>% 
  as.data.frame() %>% #default here is matrix
  `colnames<-`(c("Precipitation",
                 "stderrP",
                 "Runoff",
                 "stderrQ")) %>% #set first row to col. names
  .[-c(1,14), ] %>% #removes extra rows easily
  mutate(Month = as.numeric(row.names(.))) %>%
  mutate_all(., function(x) as.numeric(x)) #convert all columns to numeric 

```

```{r Plots, echo=FALSE}
YearPlot <-
  ggplot(data = YearData,
         aes(x = Year)) +
  geom_point(aes(y = PrecipSum, color = "Precipitation")) +
  geom_point(aes(y = RunoffSum, color = "Run off")) +
  geom_line(aes(y = PrecipSum, color = "Precipitation", group = 1)) +
  geom_line(aes(y = RunoffSum, color = "Run off", group = 1)) +
  labs(y = "Precipitation or Runoff (mm/year)",
       color = " ",
       title = "Fig. 1 Precipitation and Run Off Plotted Anually \nElbow River Watershed, AB (1993-2022)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   colour = "black"), #angling and right justified x labels
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(linewidth = 0.8)) + #making the axis thicker
  scale_colour_manual(values = c("blue4", "chocolate4")) + #Force colours
  scale_y_continuous(breaks = seq(0,1200,100)) + #best way to set tickmarks 
  scale_x_continuous(breaks = seq(1993,2022,1))

YearPlot

MonthPlot <-
  ggplot(data = MonthData,
         aes(x = Month)) +
  geom_point(aes(y = Precipitation, color = "Precipitation")) +
  geom_point(aes(y = Runoff, color = "Run off")) +
  geom_errorbar(aes(ymin = Precipitation - stderrP, 
                    ymax = Precipitation + stderrP),
                width=.2) +
  geom_line(aes(y = Precipitation, color = "Precipitation", group = 1)) +
  geom_errorbar(aes(ymin = Runoff - stderrQ,
                    ymax = Runoff + stderrQ),
                width=.3) +
  geom_line(aes(y = Runoff, color = "Run off", group = 1)) +
  labs(y = "Precipitation or Runoff (mm/month)",
       color = " ",
       title = "Fig. 2 Precipitation and Run Off Plotted Monthly \nElbow River Watershed, AB (1993-2022)") +
  theme_classic() +
  theme(axis.text.x = element_text(colour = "black"), #angling and right justified x labels
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(linewidth = 0.8)) + #making the axis thicker
  scale_colour_manual(values = c("blue4", "chocolate4")) + #Force colours
  scale_y_continuous(breaks = seq(0,140,20)) + #best way to set tickmarks
  scale_x_continuous(breaks = seq(1,12,1))

MonthPlot

```

Part two
```{r Linear Models, include=FALSE}

RainLM <-
  lm(formula = (RunoffSum ~ RainSum), 
     data = YearData)

RainLM_Summary <- summary(RainLM)

RainLM_Summary

SnowLM <- 
  lm(formula = (RunoffSum ~ SnowSum), 
     data = YearData)

SnowLM_Summary <- summary(SnowLM)

SnowLM_Summary

```

```{r Regression Plots, message=FALSE, warning=FALSE, echo=FALSE}


RainRegression <- 
  ggplot(data = YearData,
         aes(x = RainSum, y = RunoffSum)) +
  geom_smooth(method = "lm",
              se = FALSE,
              colour = "deepskyblue",
              linewidth = 0.8) +
  geom_point(colour = "blue4",
             size = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(colour = "black"), #angling and right justified x labels
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(linewidth = 0.8)) +
  labs(y = "Runoff (mm/year)",
       x = "Rain (mm/year)",
       title = "Fig. 3 Annual Rain vs. Runoff \nElbow River Watershed, AB (1993-2022)") +
  scale_x_continuous(breaks = seq(0,1000,100)) +
  scale_y_continuous(breaks = seq(0,400,50)) +
  geom_text(x = 300, y = 400, label = "Y = 58.5 + 0.41X \nR\u00B2 = 0.57 \np < 0.01")

RainRegression

SnowRegression <- 
  ggplot(data = YearData,
         aes(x = SnowSum, y = RunoffSum)) +
  geom_smooth(method = "lm",
              se = FALSE,
              colour = "deepskyblue",
              linewidth = 0.8) +
  geom_point(colour = "blue4",
             size = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(colour = "black"), #angling and right justified x labels
        axis.text.y = element_text(colour = "black"),
        axis.line = element_line(linewidth = 0.8)) +
  labs(y = "Runoff (mm/year)",
       x = "Snow (mm/year)",
       title = "Fig. 4 Annual Snow vs. Runoff \nElbow River Watershed, AB (1993-2022)") +
  scale_x_continuous(breaks = seq(0,400,50)) +
  scale_y_continuous(breaks = seq(0,400,50)) + 
  geom_text(x = 80, y = 400, label = "Y = 226.5 + 0.17X \nR\u00B2 = 0.021 \np < 0.44")

SnowRegression
```

Question 1) Precipitation is more variable, both the Variance and
Standard Error are higher for this variable. This could be from
groundwater storage acting as a buffer, as well as evapotanspiration
rates increasing with increasing precipitation.

Question 2) 
$$ 
P = ET + Q \\ 636.5 = ET + -253.5 \\ ET = 383 mm/year
$$

Question 3) 
A) 
$$ 
Y = 58.05 + 0.413(X) \\ Y = 58.05 + 0.413(473.2) \\ Y = 253.5 mm/year
$$

B)
$$ 
Y = 58.05 + 0.413(X) \\ Y = 58.05 + 0.413(50) \\ Y = 78.7 mm/year
$$

C)  

The value for part A (473 mm annually) is well within the range of data used to
build the model, while the part B value (50 mm annually) is extrapolated outside 
the range of data used. Because of the possibility for non-linearity of the 
observed values, the predicted regression should not be used to estimate 
fitted values outside the range of the measured data.

Question 4) 
A) 
These data did show a positive response between
annual rainfall and run off (R2 = 0.57, F = 37.771,28 , p = \<0.01),
contrarily there was no meaningful relationship between annual snowfall
and runoff (R2 = 0.02, F = 0.44481, 28, p = 0.45). The R2 value is a
measure of vertical or response variability from the fitted line to the
observed data, in this context it is a measure of the amount of variance
in Runoff that can be explained by either the rain or snowfall of that
year. The probability statistic, p, tells us the probability that the
response in the outcome is not a product of chance but instead there is
a significant slope, $\beta$, in the regression model. In this context a
small p-value would mean that annual precipitation is likely a factor in
the amount of run off.

B)
The above model does suggest that snowfall is unimportant to annual
runoff. This lack of correlation could be due to oversimplification in
building the model, snowfall falls over more than one calendar year
while it generally melts in a single spring. This mismatch of timing
would misplace late-year snowfall into the current year despite the
runoff it generates being formed in the following calendar year.
